The Talking Heads Experiment

### 2015-04-17 chap10/chap10.tex
cleaned up after sanity check

The \textsc{aibo} experiments were the first robotic experiments demonstrating social, situated learning in action. 
They were highly relevant for discovering the subtle interaction patterns that would work and 
could be realistically implemented in human-robot scenarios. The goal of these experiments was 
not to address the question of the origins of a 
new language in a population of agents. However, another experiment, still with the \textsc{aibo} robots,  
did substantially advance the state of the art: the Perspective Reversal Experiment carried out by Martin Loetzsch 
and Luc Steels in 2004-2005.\footnote{
Very little was published about this experiment, partly because reviewers had the greatest difficulty to grasp 
its underlying rationale or understand the technical details. The main paper is \cite{Steels:2008spatial}. 
Subsequent work by Michael Spranger significantly scaled up the experiment, as discussed in the next chapter.}

@@

Within this experimental setting three different strategies were tried in a long-term experiment where
a new strategy was injected at certain stages and performance monitored: 
\begin{itemize}
\item In stage I, agents use 
the Discrimination Game strategy and the Naming Game strategy discussed in chapters 4 and 5. A 
lexicon is emerging and there are some successful games (those for which the speaker and the hearer see the situation from 
the same perspective). But the lexicon keeps expanding and success is very limited as games where robots have a different 
perspective on the rolling ball all fail. 
\item In stage II, agents have an extended 
strategy where agents can geometrically transform a model of the scene, based on their own perspective, into a model of the scene 
as seen from the perspective of the hearer. They can then conceptualise the scene from that perspective. 
Communicative success significantly increases and the 
lexicon starts to decrease. However there is ambiguity because the hearer cannot know whether
the speaker has conceptualised the 
scene from his own viewpoint or performed a perspective reversal, leading to a large number of failed communications 
(more than 50 \%). 
\item In stage III, a more elaborate stratergy is used. The speaker now expresses whether he has performed perspective 
reversal (as in `{\itshape to your} left' vs. simply `left'). Only with this strategy are agents able to reach high 
communicative success with a stable inventory. 
\end{itemize}

\begin{figure}[p]
\centerline{
\includegraphics[width=0.75\textwidth]{chap10/figs/stages.pdf}}
\caption{\label{fig:stages}Three strategies are compared. In Stage I (left) agents categorise the event from their 
own perspective. In Stage II (middle) they conceptualise the scene from the viewpoint of the other agent as well. In Stage III (right) 
they express that they performed a perspective reversal. Only the third strategy leads to successful performance.} 
\end{figure}

>>>>

The \textsc{aibo} experiments were the first robotic experiments demonstrating social, situated learning in action. 
They were highly relevant for discovering the subtle interaction patterns that would work and 
could be realistically implemented in human-robot scenarios. The goal of these experiments was 
not to address the question of the origins of a 
new language in a population of agents. However, another experiment, still with the \textsc{aibo} robots,  
did substantially advance the state of the art: the Perspective Reversal Experiment carried out by Martin Loetzsch 
and Luc Steels in 2004--2005.\footnote{
Very little was published about this experiment, partly because reviewers had the greatest difficulty to grasp 
its underlying rationale or understand the technical details. The main paper is \cite{Steels:2008spatial}. 
Subsequent work by Michael Spranger significantly scaled up the experiment, as discussed in the next chapter.}

@@

Within this experimental setting three different strategies were tried in a long-term experiment where
a new strategy was injected at certain stages and performance monitored: 
\begin{itemize}
\item In stage I, agents use 
the Discrimination Game strategy and the Naming Game strategy discussed in chapters 4 and 5. A 
lexicon is emerging and there are some successful games (those for which the speaker and the hearer see the situation from 
the same perspective). But the lexicon keeps expanding and success is very limited as games where robots have a different 
perspective on the rolling ball all fail. 
\item In stage II, agents have an extended 
strategy where agents can geometrically transform a model of the scene, based on their own perspective, into a model of the scene 
as seen from the perspective of the hearer. They can then conceptualise the scene from that perspective. 
Communicative success significantly increases and the 
lexicon starts to decrease. However there is ambiguity because the hearer cannot know whether
the speaker has conceptualised the 
scene from his own viewpoint or performed a perspective reversal, leading to a large number of failed communications 
(more than 50 \%). 
\item In stage \textsc{iii}, a more elaborate stratergy is used. The speaker now expresses whether he has performed perspective 
reversal (as in `{\itshape to your} left' vs. simply `left'). Only with this strategy are agents able to reach high 
communicative success with a stable inventory. 
\end{itemize}

\begin{figure}[p]
\centerline{
\includegraphics[width=0.75\textwidth]{chap10/figs/stages.pdf}}
\caption{\label{fig:stages}Three strategies are compared. In Stage I (left) agents categorise the event from their 
own perspective. In Stage II (middle) they conceptualise the scene from the viewpoint of the other agent as well. In Stage \textsc{iii} (right) 
they express that they performed a perspective reversal. Only the third strategy leads to successful performance.} 
\end{figure}

### 2015-04-17 chap11/chap11.tex
cleaned up after sanity check

As a second example of how thinking in terms of semiotic networks is useful,
we turn to an experiment building further on the work on action-word learning \is{action game}
using the \textsc{aibo} reported in the previous chapter.
It was first designed and carried out with the \textsc{qrio} robot in 2007/2008 by Luc Steels and
Michael Spranger with considerable help from Martin Loetzsch\footnote{Papers describing this experiment more fully are: 
\cite{Steels:2008spatial} and \cite{Steels:2008b}.} and 
then ported in 2010 to the MYON, a robot developed by Manfred Hild and his team at the Humboldt 
University in Berlin (see \figref{fig:myon-action}) (\cite{Steels:2012b}). 
A BBC video clip of the experiment can be seen at: \url{https://youtu.be/lmoXByLkK14}.
The porting exercise showed that once we discover how to do a certain type of language game on one robot 
we can relatively easily transfer it to another one, after adapting the lowest level sensing and actuation routines.

@@

What is the nature of the constraints that we will need for the semantics of natural language? 
Take a phrase like `the ball behind the block'. It invokes the following constraints, also called 
semantic operations: 
\begin{itemize}
\item `the ball' evokes a constraint to use a prototype [BALL] to find a referent in the current scene.
\item The same holds for `the block' which uses another prototype [BLOCK]. 
\item The preposition `behind' implies another constraint, namely to use the spatial relation [BEHIND] to filter a 
set of objects (namely those found by using the [BALL] prototype) and retain only those that satisfy the behind-relation 
to an object in another set. 
\end{itemize}

@@

1. There are first {\itshape lexical constructions} that associate a word with the introduction of a semantic entity  
such as a colour category or a qualifier for indicating graded membership. So the meaning for the word `yellow' 
introduces a BIND-operation, as in: 
\begin{verbatim}
((bind colour-category ?c [yellow]))
\end{verbatim}
When the word `yellow' is encountered, a unit for this word is created with this BIND-operation as meaning and the 
word-string itself as the form. It is the unit called Yellow-unit in \figref{fig:syntactic-graded}. 
The construction also introduces a lexical category (part of speech) for the word (in this case the lexical category 
is colour-category) and specifies which variables in the meaning can be linked with other variables in other
units. The variables are typed. In this case there is one variable ?c which is of type colour-category.

>>>>

As a second example of how thinking in terms of semiotic networks is useful,
we turn to an experiment building further on the work on action-word learning \is{action game}
using the \textsc{aibo} reported in the previous chapter.
It was first designed and carried out with the \textsc{qrio} robot in 2007/2008 by Luc Steels and
Michael Spranger with considerable help from Martin Loetzsch\footnote{Papers describing this experiment more fully are: 
\cite{Steels:2008spatial} and \cite{Steels:2008b}.} and 
then ported in 2010 to the \textsc{myon}, a robot developed by Manfred Hild and his team at the Humboldt 
University in Berlin (see \figref{fig:myon-action}) (\cite{Steels:2012b}). 
A BBC video clip of the experiment can be seen at: \url{https://youtu.be/lmoXByLkK14}.
The porting exercise showed that once we discover how to do a certain type of language game on one robot 
we can relatively easily transfer it to another one, after adapting the lowest level sensing and actuation routines.

@@

What is the nature of the constraints that we will need for the semantics of natural language? 
Take a phrase like `the ball behind the block'. It invokes the following constraints, also called 
semantic operations: 
\begin{itemize}
\item `the ball' evokes a constraint to use a prototype [\textsc{ball}] to find a referent in the current scene.
\item The same holds for `the block' which uses another prototype [\textsc{block}]. 
\item The preposition `behind' implies another constraint, namely to use the spatial relation [\textsc{behind}] to filter a 
set of objects (namely those found by using the [\textsc{ball}] prototype) and retain only those that satisfy the behind-relation 
to an object in another set. 
\end{itemize}

@@

1. There are first {\itshape lexical constructions} that associate a word with the introduction of a semantic entity  
such as a colour category or a qualifier for indicating graded membership. So the meaning for the word `yellow' 
introduces a \textsc{bind}-operation, as in: 
\begin{verbatim}
((bind colour-category ?c [yellow]))
\end{verbatim}
When the word `yellow' is encountered, a unit for this word is created with this \textsc{bind}-operation as meaning and the 
word-string itself as the form. It is the unit called Yellow-unit in \figref{fig:syntactic-graded}. 
The construction also introduces a lexical category (part of speech) for the word (in this case the lexical category 
is colour-category) and specifies which variables in the meaning can be linked with other variables in other
units. The variables are typed. In this case there is one variable ?c which is of type colour-category.

### 2015-04-17 chap12/chap12.tex
cleaned up after sanity check

It is only one step now to achieve the selection and alignment of language strategies. Basically, two mechanisms needs to be 
added: \is{strategy selection}
\begin{enumerate}
\item Agents need to track the performance of their strategies by tracking the performance of the language 
systems that were built based on their strategies. In other words, not only lexical or grammatical constructions
but also strategies need to have a score. These {\itshape strategy scores} should be updated using the same 
lateral inhibition dynamics that we have been using at the level of word-meaning associations or 
more generally constructions. 
\item It is also necessary that the conceptualisations and the lexical and grammatical constructions that 
are implicated within a particular strategy are tagged with that strategy, so that if their score is updated, 
the score of the strategy can be updated as well (\figref{fig:tagging}). This is necessary  
for implementing a proper credit assignment 
mechanism and also for ensuring that a word becomes unambiguously used with a particular strategy. 
\end{enumerate}
The same word may be tagged with more than one strategy because 
a learning agent does not know which strategy has been used by the speaker and so this may lead to different
hypotheses. But in the end, agents strive for convergence so that they all use the 
same word with the same meaning. For example, a hearer cannot know whether a particular word (e.g. `yellow')
is to be interpreted using  the brightness-based or  the full-colour-based strategy, particularly
because both strategies could work in similar circumstances. For example, yellow colour chips are often the 
most bright ones and hence both strategies would work. Consequently it is unavoidable that different strategies compete
to recruit the same words, and this explains that we see this phenomenon indeed in the historically attested evolution 
of language. Indeed, many of today's hue words like `yellow', `brown' 
or `blue' were all expressing brightness-based distinctions in Old English before they became used as part of the  
Basic Colour Strategy in the late Middle English period (1350-1500).\footnote{See
\cite{MacLaury:1992} and \cite{Casson:1997}.}

@@

We will use the domain of spatial language again as a source of examples, based on experiments carried out by 
Michael Spranger. It uses the \textsc{qrio} humanoid robots 
and scenes like the one shown in \figref{fig:spatial}. An example of a constraint network from this domain 
is shown in \figref{fig:spatial-strategies}. It is the network for the phrase `near the box'. `Near' is a proximal 
relation which categorises objects in terms of how near they are to a landmark (in this case the box). In a list form the 
network looks like this: 
\begin{small}
\begin{verbatim}
1. (get-context ?ctx-1) 
2. (construct-region-proximal 
    ?region ?ctx-1 ?landmark ?cat) 
3. (bind proximal-category ?cat NEAR) 
4. (apply-selection ?landmark ?landmarks ?selector-2)
5. (apply-class ?landmarks ?ctx ?landmark-class) 
6. (bind selector ?selector-2 UNIQUE) 
7. (get-context ?ctx) 
8. (bind object-class ?landmark-class BOX)
\end{verbatim}
\end{small}
The order of these operations does not matter as values propagate and constraints become active until no further progress
can be made. Operations in 4, 5, 6, 7 and 8 provide the referent of `the box'. The following contributions 
can be made by each of the cognitive operations: The predicate BOX gets bound to ?landmark-class in operation 8. Object-class is 
the type of ?landmark-class. The context, i.e. all objects in the present situation, gets bound to ?ctx in operation 7. 
The selector UNIQUE gets bound to ?selector-2 in operation 6. 
Then the operation apply-class is applied to all the objects in ?ctx (the context) to find the possible landmarks which 
get bound to ?landmarks using the ?landmark-class (BOX). This set is assumed to be a singleton and so 
in operation 4 the unique element in the set of ?landmarks can be selected 
and bound to ?landmark. 
\enlargethispage{1\baselineskip}
The remaining three operations apply the proximal relation. In operation 3 ?cat is bound to the proximal-category NEAR. 
In operation 1, ?ctx-1 gets bound to 
the context. And finally in operation 2, the region around the landmark, which satisfies the category ?cat (NEAR), can 
be computed and bound to ?region by ``construct-region-proximal''. The construction-region-proximal constraint has to perform first 
a geometric transformation on the scene, constructing a world model from the viewpoint of the landmark, i.e. the box.

@@

This section reported on experiments in the 2010-2015 time period that have tried to scale
up language dynamics from a single strategy (as in the Talking Heads) to 
multiple strategies. This lead to the discovery that we need a dual dynamics: on the level of language systems
governed by one or more strategies, and on the level of language strategies. We have applied a cultural Darwinian dynamics 
on both levels. How that works for language systems is quite clear now, but many questions remain and much 
more work needs to be done to see how it would work on the level of language strategies. This chapter reported  
some of the progress made recently. It introduced three steps: niche analysis as a way to understand the evolutionary landscape, 
selection and alignment of language strategies, and the generation of new language strategies.

>>>>

It is only one step now to achieve the selection and alignment of language strategies. Basically, two mechanisms needs to be 
added: \is{strategy selection}
\begin{enumerate}
\item Agents need to track the performance of their strategies by tracking the performance of the language 
systems that were built based on their strategies. In other words, not only lexical or grammatical constructions
but also strategies need to have a score. These {\itshape strategy scores} should be updated using the same 
lateral inhibition dynamics that we have been using at the level of word-meaning associations or 
more generally constructions. 
\item It is also necessary that the conceptualisations and the lexical and grammatical constructions that 
are implicated within a particular strategy are tagged with that strategy, so that if their score is updated, 
the score of the strategy can be updated as well (\figref{fig:tagging}). This is necessary  
for implementing a proper credit assignment 
mechanism and also for ensuring that a word becomes unambiguously used with a particular strategy. 
\end{enumerate}
The same word may be tagged with more than one strategy because 
a learning agent does not know which strategy has been used by the speaker and so this may lead to different
hypotheses. But in the end, agents strive for convergence so that they all use the 
same word with the same meaning. For example, a hearer cannot know whether a particular word (e.g. `yellow')
is to be interpreted using  the brightness-based or  the full-colour-based strategy, particularly
because both strategies could work in similar circumstances. For example, yellow colour chips are often the 
most bright ones and hence both strategies would work. Consequently it is unavoidable that different strategies compete
to recruit the same words, and this explains that we see this phenomenon indeed in the historically attested evolution 
of language. Indeed, many of today's hue words like `yellow', `brown' 
or `blue' were all expressing brightness-based distinctions in Old English before they became used as part of the  
Basic Colour Strategy in the late Middle English period (1350--1500).\footnote{See
\cite{MacLaury:1992} and \cite{Casson:1997}.}

@@

We will use the domain of spatial language again as a source of examples, based on experiments carried out by 
Michael Spranger. It uses the \textsc{qrio} humanoid robots 
and scenes like the one shown in \figref{fig:spatial}. An example of a constraint network from this domain 
is shown in \figref{fig:spatial-strategies}. It is the network for the phrase `near the box'. `Near' is a proximal 
relation which categorises objects in terms of how near they are to a landmark (in this case the box). In a list form the 
network looks like this: 
\begin{small}
\begin{verbatim}
1. (get-context ?ctx-1) 
2. (construct-region-proximal 
    ?region ?ctx-1 ?landmark ?cat) 
3. (bind proximal-category ?cat NEAR) 
4. (apply-selection ?landmark ?landmarks ?selector-2)
5. (apply-class ?landmarks ?ctx ?landmark-class) 
6. (bind selector ?selector-2 UNIQUE) 
7. (get-context ?ctx) 
8. (bind object-class ?landmark-class BOX)
\end{verbatim}
\end{small}
The order of these operations does not matter as values propagate and constraints become active until no further progress
can be made. Operations in 4, 5, 6, 7 and 8 provide the referent of `the box'. The following contributions 
can be made by each of the cognitive operations: The predicate \textsc{box} gets bound to ?landmark-class in operation 8. Object-class is 
the type of ?landmark-class. The context, i.e. all objects in the present situation, gets bound to ?ctx in operation 7. 
The selector \textsc{unique} gets bound to ?selector-2 in operation 6. 
Then the operation apply-class is applied to all the objects in ?ctx (the context) to find the possible landmarks which 
get bound to ?landmarks using the ?landmark-class (\textsc{box}). This set is assumed to be a singleton and so 
in operation 4 the unique element in the set of ?landmarks can be selected 
and bound to ?landmark. 
\enlargethispage{1\baselineskip}
The remaining three operations apply the proximal relation. In operation 3 ?cat is bound to the proximal-category \textsc{near}. 
In operation 1, ?ctx-1 gets bound to 
the context. And finally in operation 2, the region around the landmark, which satisfies the category ?cat (\textsc{near}), can 
be computed and bound to ?region by ``construct-region-proximal''. The construction-region-proximal constraint has to perform first 
a geometric transformation on the scene, constructing a world model from the viewpoint of the landmark, i.e. the box.

@@

This section reported on experiments in the 2010--2015 time period that have tried to scale
up language dynamics from a single strategy (as in the Talking Heads) to 
multiple strategies. This lead to the discovery that we need a dual dynamics: on the level of language systems
governed by one or more strategies, and on the level of language strategies. We have applied a cultural Darwinian dynamics 
on both levels. How that works for language systems is quite clear now, but many questions remain and much 
more work needs to be done to see how it would work on the level of language strategies. This chapter reported  
some of the progress made recently. It introduced three steps: niche analysis as a way to understand the evolutionary landscape, 
selection and alignment of language strategies, and the generation of new language strategies.

### 2015-04-17 chap8/chap8.tex
cleaned up after sanity check

A total of 8000 words and 500 concepts were created by the agents, with a core vocabulary consisting of 100 basic words expressing concepts like up, down, left, right, green, red, large, small, etc. Of these, 8 words represent a large majority of words used (about 80 \%). 4 of these words refer to the position of objects: GOREWA (top), \textsc{down} (bottom), WOGGLESPLAT (left), and SESUBIPU (right). 4 other words refer to colors: ROUGE (red), KAZOZO (green), WEGIRIRA (blue), and EMPTY (light). The distribution of these words after 130,000 games is shown in \figref{fig:words-used}.

@@

\figref{fig:woggle} shows the semiotic dynamics related to an expression of the meaning for the 
concept `left', i.e. a horizontal position 
between 0.0 and 0.25 (after scaling). The word "wogglesplat" becomes dominant
although there is a very strong competition in the beginning. 
We see that users try to give other words to the same meaning, such as "gauche" or "links" (both words expressing left 
in French and German or Dutch respectively). We also see that other words such as "red" or "yellow" get associated with this 
meaning, because the hearer may guess the wrong meaning in learning the word for left.

>>>>

A total of 8000 words and 500 concepts were created by the agents, with a core vocabulary consisting of 100 basic words expressing concepts like up, down, left, right, green, red, large, small, etc. Of these, 8 words represent a large majority of words used (about 80 \%). 4 of these words refer to the position of objects: \textsc{gorewa} (top), \textsc{down} (bottom), \textsc{wogglesplat} (left), and \textsc{sesubipu} (right). 4 other words refer to colors: \textsc{rouge} (red), \textsc{kazozo} (green), \textsc{wegirira} (blue), and \textsc{empty} (light). The distribution of these words after 130,000 games is shown in \figref{fig:words-used}.

@@

\figref{fig:woggle} shows the semiotic dynamics related to an expression of the meaning for the 
concept `left', i.e. a horizontal position 
between 0.0 and 0.25 (after scaling). The word ``wogglesplat" becomes dominant
although there is a very strong competition in the beginning. 
We see that users try to give other words to the same meaning, such as "gauche" or "links" (both words expressing left 
in French and German or Dutch respectively). We also see that other words such as "red" or "yellow" get associated with this 
meaning, because the hearer may guess the wrong meaning in learning the word for left.
